The Gender Pay Gap in the General Social Survey

### 2023-08-28 Paper/paper.Rmd
using log income everywhere

---
title: "The Gender Pay Gap in the General Social Survey"
author: "Alex Amaguaya, Lea Bergmann"
cols_authors: 3
subtitle: "Statistical Learning"
deadline: "tomorrow"
type: "Term Paper"
date: "today"
supervisors: "Prof. Dr. Thomas Deckers, Karolina Gliszczynska"
output:
  pdf_document:
    keep_tex: yes
    template: template.tex
    fig_caption: yes
    citation_package: biblatex
    number_sections: true
toc: true
lot: true
lof: true
graphics: true
biblio-title: References
fontsize: 11pt
geometry: lmargin=2.5cm,rmargin=2.5cm,tmargin=2.5cm,bmargin=2.5cm
biblio-files: references.bib
classoption: a4paper
---

@@

[Write something about cross validation?]

@@

-   **year**: survey year
-   **realrinc**: respondent's base income (in constant 1986 USD)
-   **age**: respondent's age in years
-   **occ10**: respondent's occupation code
-   **occrecode**: recode of the occupation code into one of 11 main categories
-   **prestg10**: respondent's occupational prestige score
-   **childs**: number of children
-   **wrkstat**: work status of the respondent (full-time, part-time, temporarily not working, unemployed (laid off), etc.)
-   **gender**: respondent's gender
-   **educcat**: respondent's degree level (Less Than High School, High School, etc.)
-   **maritalcat**: respondent's marital status (Married, Widowed, Divorced, etc.)

@@

![table](includes/reg_table.png){width="317"}

@@

![matrix_corr](includes/corr_matrix.png)

@@

![feature_process](includes/feature_process.png)

@@

Given some extreme outliers in the income distribution of our data (which is typical for wage data) we follow the economic literature [@Ermini] and fit a logistic model to our data, using the log of `realrinc` as our dependent variable. This results in a higher RMSE than the linear model (25,909.82).

@@

$$
RSS + \lambda \sum^p _{j=1} \beta^2_j 
$$

@@

``` r
ridge <- train(formula, data = train,
  method = "glmnet", trControl = trainControl(method = "cv", number = 3),
  tuneGrid = expand.grid(alpha = 0, lambda = seq(0, 500, 1))) 
```

Perform a ridge regression to predict real incomes yields an RMSE of 25,524.85.

@@

``` r
lasso <- train(formula, data = train, 
       method = "glmnet", trControl = trainControl(method = "cv"), 
       tuneGrid = expand.grid(alpha = 1, lambda = seq(0,500,1)))
```

To perform the lasso regression in `R` we set $\alpha$ equal to one. The lasso regression yields a lower RMSE than the ridge regression (25,504.89 vs. 25,524.85) and even a lower RMSE than the linear regression.

In a final step we use elastic net regularization which is a linear combination of ridge and lasso regression. The elastic net regularization allows $\alpha$ to vary between zero and one. This results in an RMSE of 25,504.18.

@@

According to the literature review, tree-based models perform considerably well in range in predicting income ranges (classification problems). For this reason, the Random forest algorithm is used for modeling. This experiment use the income without logarithm transformation, the initial features and some interaction variables.

@@

Due to the number of hyper parameters in Random Forest, we use the `lightgbm` and `mlr3` packages to model and tune the hyper parameter in a time-optimal way. After many attempts, we found the best hyper parameters that produced the lowest error. The hyper parameters tuned were `max_depth`, `num_leaves`, `num_iterations` and others. In addition, we used a random search approach in the tuning process with a cross-validation of 3.
Once we found the model along with the hyper parameters with lowest error in the cross-validation, we proceeded to evaluate this model on the test data set and obtained a RMSE of 31745.54. The hyper parameters of the best model are shown in the following table.

@@

In addition, we can analyze the importance of the features within the modeling process using Random Forest, and these results are shown in the following table.

@@

Gradient Boosting also has many hyper parameters like Random Forest and we decided to apply a similar approach to the previous algorithm to find the best model with the hyper parameters (random search approach in the tuning process with a cross-validation of 3). Furthermore, the experiments with Gradient Boosting used the target variable without logarithm transformation.

@@

We tried many experiments and found the best hyper parameters that produced the lowest error. In Gradient Boosting modeling we adjusted the hyper parameters such as `max_depth`, `num_leaves`, `min_data_in_leaf`, `feature_fraction` and others.
Once the best model was found, it obtained an RMSE of 29207.55 on the test data set. The hyper parameters of the best model are shown in the following table.

@@

These results show that in tree-based methods, the Gradient Boosting performance overcomes Random Forest, in other words, the error of Gradient Boosting is approximately 8% lower than the error of Random Forest. In addition, the error of these model are large than the linear models.

>>>>

---
title: "The Gender Pay Gap in the General Social Survey"
author: "Alex Amaguaya, Lea Bergmann"
cols_authors: 3
subtitle: "Statistical Learning"
deadline: "tomorrow"
type: "Term Paper"
date: "today"
supervisors: "Prof. Dr. Thomas Deckers, Karolina Gliszczynska"
output:
  pdf_document:
    keep_tex: yes
    template: template.tex
    fig_caption: yes
    citation_package: biblatex
    number_sections: true
toc: true
lot: true
lof: true
graphics: true
biblio-title: References
fontsize: 11pt
geometry: lmargin=2.5cm,rmargin=2.5cm,tmargin=2.5cm,bmargin=2.5cm
biblio-files: ref.bib
classoption: a4paper
---

@@

Furthermore, we use cross validation (CV) to assess the performance of a model and to select appropriate hyperparameters. CV prevents overfitting and helps in model selection.

@@

-   **year**: survey year

-   **realrinc**: respondent's base income (in constant 1986 USD)

-   **age**: respondent's age in years

-   **occ10**: respondent's occupation code

-   **occrecode**: recode of the occupation code into one of 11 main categories

-   **prestg10**: respondent's occupational prestige score

-   **childs**: number of children

-   **wrkstat**: work status of the respondent (full-time, part-time, temporarily not working, unemployed (laid off), etc.)

-   **gender**: respondent's gender

-   **educcat**: respondent's degree level (Less Than High School, High School, etc.)

-   **maritalcat**: respondent's marital status (Married, Widowed, Divorced, etc.)

@@

![Linear Model: Inclusion of Interaction Variables](includes/reg_table.png){width="317"}

@@

![Correlation Matrix](includes/corr_matrix.png)

@@

![Feature Engineering](includes/feature_process.png)

@@

Given some extreme outliers in the income distribution of our data (which is typical for wage data) we follow the economic literature [@Ermini] and fit a logistic model to our data, using the log of `realrinc` as our dependent variable. This results in a higher RMSE than the linear model (25,909.82). Still, the linear model yields predictions of `realrinc` that are negative, what is implausible. Using the log of `realrinc` only positive predictions are generated. Therefore, we prefer using the log for the following predictions.

@@

$$
y = RSS + \lambda \sum^p _{j=1} \beta^2_j 
$$

@@

``` r
ridge <- train(formula, data = train,
  method = "glmnet", trControl = trainControl(method = "cv", number = 3),
  tuneGrid = expand.grid(alpha = 0, lambda = seq(0,1,0.001)))
```

Perform a ridge regression to predict real incomes yields an RMSE of 26,062.58.

@@

``` r
lasso <- train(formula, data = train, 
       method = "glmnet", trControl = trainControl(method = "cv"), 
       tuneGrid = expand.grid(alpha = 1, lambda = seq(0,1,0.001)))
```

To perform the lasso regression in `R` we set $\alpha$ equal to one. The lasso regression yields a lower RMSE than the ridge regression (25,918.25 vs. 26,062.58).

In a final step we use elastic net regularization which is a linear combination of ridge and lasso regression. The elastic net regularization allows $\alpha$ to vary between zero and one. This results in an RMSE of 25,918.81.

@@

According to the literature review, tree-based models perform considerably well in range in predicting income ranges (classification problems). For this reason, we use the random forest algorithm for modeling. This experiment use the income without logarithm transformation, the initial features and some interaction variables.

@@

Due to the number of hyper parameters in random forest, we use the `lightgbm` and `mlr3` packages to model and tune the hyper parameter in a time-optimal way. After many attempts, we found the best hyper parameters that produced the lowest error. The hyper parameters tuned were `max_depth`, `num_leaves`, `num_iterations` and others. In addition, we used a random search approach in the tuning process with a cross-validation of 3. Once we found the model along with the hyper parameters with lowest error in the cross-validation, we proceeded to evaluate this model on the test data set and obtained a RMSE of 31,745.54. The hyper parameters of the best model are shown in the following table.

@@

In addition, we can analyze the importance of the features within the modeling process using random forest, and these results are shown in the following table.

@@

Gradient Boosting also has many hyper parameters like random forest and we decided to apply a similar approach to the previous algorithm to find the best model with the hyper parameters (random search approach in the tuning process with a cross-validation of 3). Furthermore, the experiments with Gradient Boosting use the target variable without logarithm transformation.

@@

We tried many experiments and found the best hyper parameters that produced the lowest error. In Gradient Boosting modeling we adjusted the hyper parameters such as `max_depth`, `num_leaves`, `min_data_in_leaf`, `feature_fraction` and others. Once the best model was found, it obtained an RMSE of 29,207.55 on the test dataset. The hyper parameters of the best model are shown in the following table.

@@

These results show that in tree-based methods, the Gradient Boosting performance overcomes Random Forest, in other words, the error of Gradient Boosting is approximately 8% lower than the error of Random Forest. In addition, the errors of these models are larger than the linear models.
